---
title: "Analysis"
output: pdf_document
---

```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library(corrplot)
library(gridExtra)
library(car)
```


# Data Cleaning and Preprocessing
```{r}
df <- read.csv("data/data.csv")
head(df)
```

## Basic Cleaning of the Data
```{r}
# Adding AVG_S11
df <- df %>% 
  mutate("AVG_S11" = (MAT_S11 + CR_S11 + CC_S11 + BIO_S11 + ENG_S11) / 5)

# Making continuous columns into factors
continuous_cols <- c("MAT_S11", "CR_S11", "CC_S11", "BIO_S11", "ENG_S11", "G_SC", "AVG_S11",
                     "QR_PRO", "CR_PRO", "CC_PRO", "ENG_PRO", "WC_PRO", "FEP_PRO", "G_SC")

df <- df%>%
  mutate(across(!all_of(continuous_cols), as.factor))

# Checking missing observations
print("Missing Observations in Each Column")
colSums(is.na(df))

# removing identifies and variables derived from others
df <- df %>% select(-c(...10, X, COD_S11, Cod_SPro, QUARTILE, X2ND_DECILE, PERCENTILE))
```

## Investigating Variables
```{r}
# Separating Numeric and Categorical
df_numeric <- df %>% select(all_of(continuous_cols))
df_categorical <- df %>% select(-c(continuous_cols))

# Checking summary
summary(df_numeric)
summary(df_categorical) # frequency of all values
```

## Cleaning and Processing: Categorical Data
```{r}
# Making each value in each category unique 
lapply(df_categorical, function(x) head(unique(x), 15))

# Getting number of levels for each categorical variable
df_categorical %>% summarise_all(~ n_distinct(.))

# Getting number of missing values (encoded as 0's)
df_categorical %>%
  summarise_all(~ sum(. == 0))
```

```{r}
# since not too many missing values, we can just drop them
df <- df %>%
  filter(EDU_FATHER != 0, EDU_MOTHER != 0, OCC_FATHER != 0, OCC_MOTHER != 0, 
         STRATUM != 0, SISBEN != 0, REVENUE != 0, JOB != 0)

# selecting again for future but with the cleaned data
df_numeric <- df %>% select(all_of(continuous_cols))
df_categorical <- df %>% select(-c(continuous_cols))
```

# Visualizations

## Numerical Variables: Visualizations
```{r}
# Plots to show distributions of each variable
plots <- lapply(names(df_numeric), function(feature) {
  ggplot(df_numeric, aes_string(x = feature)) + geom_histogram(bins = 30) + labs(title = paste("Distribution of ", feature)) +
    theme_minimal() + theme(plot.title = element_text(size = 8)) + theme(axis.title.x = element_text(size=8))
})

grid.arrange(grobs = plots, ncol = 4)
```

## Categorical Features: Transformations
```{r}
# Label Encoding Categorical Variables as numbers
# For ordinal Categories (EDU_FATHER, EDU_MOTHER, PEOPLE_HOUSE, REVENUE, STRATUM, SISBEN)
# SEL and SEL_IHE are already encoded as ordinal

ordinal_features <- list(
  EDU_FATHER = c("Not sure", "Ninguno", "Incomplete primary", "Complete primary", 
                 "Incomplete Secundary", "Complete Secundary", 
                 "Incomplete technical or technological", "Complete technique or technology",
                 "Incomplete Professional Education", "Complete professional education",
                 "Postgraduate education"),
  EDU_MOTHER = c("Not sure", "Ninguno", "Incomplete primary", "Complete primary", 
                 "Incomplete Secundary", "Complete Secundary", 
                 "Incomplete technical or technological", "Complete technique or technology",
                 "Incomplete Professional Education", "Complete professional education",
                 "Postgraduate education"),
  PEOPLE_HOUSE = c("One", "Two", "Three", "Four", "Five", "Six", "Seven", "Eight", "Nueve",
                   "Ten", "Once", "Twelve or more"),
  STRATUM = c("Stratum 1", "Stratum 2", "Stratum 3", "Stratum 4", "Stratum 5", "Stratum 6"),
  SISBEN = c("It is not classified by the SISBEN", "Level 1", "Level 2", "Level 3",
             "Esta clasificada en otro Level del SISBEN"),
  JOB = c("No", "Yes, 20 hours or more per week", "Yes, less than 20 hours per week")
)

df_cat_encoded <- data.frame(df_categorical)
df_cat_encoded[names(ordinal_features)] <- lapply(names(ordinal_features), function(feature){
  df_cat_encoded[[feature]] <- as.factor(as.integer(factor(df_categorical[[feature]], levels = ordinal_features[[feature]], ordered = TRUE)))
})
```

```{r}
# special ordinal encoding for Revenue and people_house since can tell 
# value of difference from one category to another
# revenue level is average of LMMW
df_cat_encoded <- df_cat_encoded %>%
  mutate(REVENUE = case_when(
    REVENUE == "less than 1 LMMW" ~ 0.5,
    REVENUE == "Between 1 and less than 2 LMMW" ~ 1.5,
    REVENUE == "Between 2 and less than 3 LMMW" ~ 2.3, 
    REVENUE == "Between 3 and less than 5 LMMW" ~ 4, 
    REVENUE == "Between 5 and less than 7 LMMW" ~ 6, 
    REVENUE == "Between 7 and less than 10 LMMW" ~ 8.5,
    REVENUE == "10 or more LMMW" ~ 10
  ))

df_cat_encoded$PEOPLE_HOUSE <- as.numeric(df_cat_encoded$PEOPLE_HOUSE)
```

```{r}
# Encoding All the Yes No Variables
# Yes no variables have inconsistent encoding in original dataset (ie some have Yes = 1 No = 0, others have No = 1, Yes = 0)
yes_no_encoding <- c("No", "Yes") # encodes all no's as 0 and yes as 1
yes_no_features <- c("INTERNET", "TV", "COMPUTER", "WASHING_MCH", "MIC_OVEN", 
                        "CAR", "FRESH", "PHONE", "MOBILE", "DVD")
df_cat_encoded[yes_no_features] <- lapply(yes_no_features, function(feature){
  df_cat_encoded[[feature]] <- as.factor(as.integer(factor(df_categorical[[feature]], levels = yes_no_encoding)) - 1)
})
```

```{r}
# gender encoding unique for interpretation, make it yes no for male
df_cat_encoded$GENDER <- as.factor(as.integer(factor(df_categorical$GENDER)) - 1)
names(df_cat_encoded)[names(df_cat_encoded) == "GENDER"] <- "GENDER_MALE"
```

```{r}
# encoding everything else
remaining_features <- c("SCHOOL_NAME", "SCHOOL_NAT", "SCHOOL_TYPE", "UNIVERSITY",
                        "ACADEMIC_PROGRAM", "SEL", "SEL_IHE", "OCC_FATHER", "OCC_MOTHER")

df_cat_encoded[remaining_features] <- lapply(remaining_features, function(feature){
  df_cat_encoded[[feature]] <- as.factor(as.integer(factor(df_categorical[[feature]])))
})
```

```{r}
# final category encoded dataframe
head(df_cat_encoded)
```

## Categorical Variables: Visualizations
```{r}
# visualizing categorical features with less than 12 unique observations per category
categorical_features <- names(df_cat_encoded)
df_cat_lt_12 <- df_cat_encoded[sapply(df_cat_encoded[categorical_features], function(x) length(unique(x)) < 12)]

plots <- lapply(names(df_cat_lt_12)[0:16], function(feature) {
  ggplot(df_cat_encoded, aes_string(x = feature)) + geom_histogram(stat="count") + labs(title = paste("Distribution of ", feature)) +
    theme_minimal() + theme(plot.title = element_text(size = 8)) + theme(axis.title.x = element_text(size=8))
})

grid.arrange(grobs = plots, ncol = 4)

plots <- lapply(names(df_cat_lt_12)[16:length(names(df_cat_lt_12))], function(feature) {
  ggplot(df_cat_encoded, aes_string(x = feature)) + geom_histogram(stat="count") + labs(title = paste("Distribution of ", feature)) +
    theme_minimal() + theme(plot.title = element_text(size = 8)) + theme(axis.title.x = element_text(size=8))
})

grid.arrange(grobs = plots, ncol = 4)
```

```{r}
# TODO: FIX ENCODINGS FOR THIS GARBAGE
# Plots for All Remaining Categorical Features
df_cat_gt_12 <- df_cat_encoded[sapply(df_cat_encoded[categorical_features], function(x) length(unique(x)) >= 12)]

for (feature in names(df_cat_gt_12)) {
  print(
    ggplot(df_cat_encoded, aes_string(x = feature)) + geom_histogram(stat="count") + 
    labs(title = paste("Distribution of ", feature)) + theme_minimal() + 
    theme(plot.title = element_text(size = 8)) + theme(axis.title.x = element_text(size=8))
  )
}

# grid.arrange(grobs = plots, ncol = 4)
```

## Fully Cleaned and Processed Dataset
```{r}
df_encoded <- cbind(df_cat_encoded, df_numeric)
```

# Exploratory Data Analysis
## Numerical Features Only

```{r}
# dropping features that are from same test period as outcome
numeric_to_drop <- c("QR_PRO", "CR_PRO", "CC_PRO", "ENG_PRO", "WC_PRO", "FEP_PRO")
df_numeric_filtered <- df_numeric %>% select(-all_of(numeric_to_drop))
```


```{r}
# Scatterplots
plot(df_numeric_filtered)
```


```{r}
# Correlation matrix heatmap
corr_matrix <- cor(df_numeric_filtered)
corrplot(corr_matrix, method = "color", tl.srt = 45, addCoef.col = "white", number.cex = 0.8)
```

## Numerical and Categorical
```{r}
# Boxplots for each numerical and categorical feature with less than 13 unique values
for (cat_feature in names(df_cat_lt_12)) {
  plots <- lapply(names(df_numeric), function(numeric_feature) {
    ggplot(df_encoded, aes_string(x = cat_feature, y = numeric_feature)) +
    geom_boxplot() +
    labs(title = paste(numeric_feature, " and ", cat_feature), y = numeric_feature,
         x = cat_feature) +
    theme_minimal() +
    theme(plot.title = element_text(size = 8)) +
    theme(axis.title.x = element_text(size=8))
  })

  grid.arrange(grobs = plots, ncol = 4)
}
```

# Creating Models
```{r}
# TODO: CLEAN UP ALL CODE AFTER THIS

# Dropping a few columns

# AVG_S11 dropped to prevent perfect multicolinearity
# PRO tests dropped since G_SC calculated from PRO tests
# Other indicators such as decile and percentile dropped as not informative
# Any categorical variable with 20+ category dropped because unusable
features <- c(names(df_numeric), names(df_cat_encoded))
features <- features[!features %in% c(
  "QR_PRO", "CR_PRO", "CC_PRO", "ENG_PRO", "WC_PRO", "FEP_PRO",
  "PERCENTILE", "G_SC", "SCHOOL_NAME", "UNIVERSITY", 
  "ACADEMIC_PROGRAM", "X2ND_DECILE", "QUARTILE",
  "AVG_S11"
  )]
```

## Helper Functions for Creating and Assessing Models

### Feature Selection
```{r}
# feature_selection
feature_selection <- function(model, features_list, threshold, excluded = c(), data = df_encoded) {
  
  features_not_in_model <- setdiff(features_list, excluded)
  rss_old <- deviance(model)
  rss_reduction_threshold <- threshold 
  rss_reduction <- Inf
  model <- model
  while (rss_reduction >= rss_reduction_threshold) {
    rss_values <- numeric(length(features_not_in_model))
    for (i in 1:length(features_not_in_model)) {
      new_formula <- as.formula(paste(Reduce(paste, deparse(formula(model))), " + ", features_not_in_model[i]))
      model_temp <- lm(new_formula, data = data)
      rss_new <- deviance(model_temp)
      rss_reduction_temp <- rss_old - rss_new
      rss_values[i] <- rss_reduction_temp  
    }
    max_rss_ind <- which.max(rss_values)
    new_formula <- as.formula(paste(Reduce(paste, deparse(formula(model))), " + ", features_not_in_model[max_rss_ind]))
    model <- lm(new_formula, data = data)
    rss_reduction <- max(rss_values)
    rss_old <- deviance(model)
    features_not_in_model <- features_not_in_model[-max_rss_ind]
    print(rss_reduction)
  }
  return(model)
}
```

```{r}
# function for finding the first predictor to add
feature_selection_first <- function(model, feature_list, df = df_encoded) {
  rss_values <- numeric(length(feature_list))
  # Getting first ele
  for (i in 1:length(feature_list)){
    new_formula <- as.formula(paste(Reduce(paste, deparse(formula(model))), " + ", features_not_in_model[i]))
    model_temp <- lm(new_formula, data = df)
    rss_new <- deviance(model_temp)
    rss_values[i] <- rss_new
  }
  min_rss_ind <- which.min(rss_values)
  print(min_rss_ind)
  print(feature_list[i])
  new_formula <- as.formula(paste(Reduce(paste, deparse(formula(model))), " + ", features_not_in_model[min_rss_ind]))
  model <- lm(new_formula, data = df)
  return(model)
}
```

### Plot Diagnostics
```{r}
# model diagnostics function to create residual plots
plot_diagnostics <- function(model, data = df_encoded) {
 res <- residuals(model)
  y_pred <- fitted(model)
  plot(y_pred, res, abline(h = 0, lty = 2, col = "red"))
  
  std_residual <- rstandard(model)
  hist(std_residual, breaks = 50, xlab="Standardized Residual", main="Histogram of Standardized Residuals")
  
  qqnorm(res)
  
  y_pred <- fitted(model)
  plot(y_pred, data$G_SC)

}
```

### Adding Interaction Terms
```{r}
interactions <- c()
for (i in 1:(length(features)-1)) {
  for (j in (i+1):length(features)) {
    interaction <- paste(c(features[i], features[j]), collapse = ":")
    interactions <- append(interactions, interaction)
  }
}

features_with_int <- c(features, interactions)
```

## Model Prev: Preliminary Model 
```{r}
# preliminary model for comparison
model_prev <- lm(G_SC~ AVG_S11 + SEL+ SEL_IHE + GENDER_MALE + SCHOOL_NAT, data=df_encoded)
summary(model_prev)
anova(model_prev)
plot_diagnostics(model_prev)
```

## Model 0: Baseline Model
```{r}
# baseline model with all features
formula <- as.formula(paste("G_SC ~", paste(features, collapse = " + ")))
model0 <- lm(as.formula(formula), data = df_encoded)

summary(model0)
vif(model0)                # assessing multicolinearity
plot_diagnostics(model0)   # diagnostics for baseline model
```

## Model 1: Naive Feature Selection

Trying feature selection with only raw variables, no transformations,
no interactions.

```{r}
# SEL_IHE was selected by feature_selection_first
model1 <- lm("G_SC ~ SEL_IHE", data = df_encoded)  
model1 <- feature_selection(model1, features, 4000, c("SEL_IHE"))

summary(model1)
anova(model1)
vif(model1)
deviance(model1)
plot_diagnostics(model1)
```

## Model 2: Just Interactions

Trying feature selection with only interaction terms.
```{r}
model2 <- lm("G_SC ~ SEL_IHE", data = df_encoded)
model2 <- feature_selection(model2, features_with_int, 80000, c("SEL_IHE"))

summary(model2)
anova(model2)
plot_diagnostics(model2)
```

## Model 3: Adding Interactions to Model 1
```{r}
# taking original model and adding interactions
model1_preds <- attr(terms(model1), "term.labels")
model3 <- feature_selection(model1, features_with_int, 30000, excluded=model1_preds)

summary(model3)
anova(model3)
plot_diagnostics(model3)
```

## Final Model Contenders
```{r}
# best model without interactions
model4 <- lm(G_SC ~ MAT_S11 + ENG_S11 + CR_S11 + CC_S11 + BIO_S11 + SEL_IHE + OCC_FATHER, data=df_encoded)
summary(model4)
anova(model4)
plot_diagnostics(model4)
deviance(model4)
vif(model4)
```

```{r}
# best model with interactions
model5 <- lm(G_SC ~ MAT_S11 + ENG_S11 + CR_S11 + CC_S11 + BIO_S11 + SEL_IHE + OCC_FATHER + I(MAT_S11*BIO_S11), data=df_encoded)
summary(model5)
anova(model5)
plot_diagnostics(model5)
deviance(model5)
vif(model5)
```

We chose to go ahead with model 5

```{r}
# transfering data into .dta format for stata regression tables
library(haven)

df$REVENUE <- as.numeric(df_encoded$REVENUE)

write_dta(df, "df_stata.dta")
```

# Problematic Observations

## Outliers
```{r}
# 3 outliers
stanres <- rstudent(model5)
sum(stanres >= 4)
outlier_indices <- which(stanres >= 4)

df_outlier <- df[outlier_indices, ]
df_outlier
```
All individuals who are outliers did really poorly on S11 (avg of ~40%) but did moderately better on G_SC (avg of ~60%)
- Score for the PRO exams on average were just a lot higher than on S11 exams. Although can't see from the summary of the numeric data, from visualizations, the pro tests are a lot more left skewed. But should still be noted that G_SC and FEP_PRO are fairly normally distributed.
- Could be like due to difference in times resulting in the student having more time to reevaluate decisions and study for pro tests?

## Leverage Points
```{r}
# Leverage Points
# 124 leverage points
# TODO: 114 leverage points when using p+1 instead of just p
# ^not todo ur not allowed to use p+1 here. the available options 
# are 2(p+1)/n or 3p/n there's no in-between lmao
leverage_values <- hatvalues(model5)
p <- length(coef(model5)) - 1
n <- nrow(df_encoded) 
leverage_threshold <- 3*(p+1)/n       # not using 2*(p+1) / n, being more strict

leverage_indices <- which(leverage_values > leverage_threshold)
df_leverage <- df_encoded[leverage_indices, ]
head(df_leverage)
```

### Visualizations for further testing LPs
```{r}
# Visualizations of Numeric Data for Leverage Points
plots <- lapply(names(df_numeric), function(feature) {
  ggplot(df_leverage, aes_string(x = feature)) + geom_histogram(bins = 30) + labs(title = paste("Distribution of ", feature)) +
    theme_minimal() + theme(plot.title = element_text(size = 8)) + theme(axis.title.x = element_text(size=8))
})

grid.arrange(grobs = plots, ncol = 4)
```

```{r}
# visualizations for categorical features of leverage points data
plots <- lapply(names(df_cat_lt_12)[0:16], function(feature) {
  ggplot(df_leverage, aes_string(x = feature)) + geom_histogram(stat="count") + labs(title = paste("Distribution of ", feature)) +
    theme_minimal() + theme(plot.title = element_text(size = 8)) + theme(axis.title.x = element_text(size=8))
})

grid.arrange(grobs = plots, ncol = 4)

plots <- lapply(names(df_cat_lt_12)[16:length(names(df_cat_lt_12))], function(feature) {
  ggplot(df_leverage, aes_string(x = feature)) + geom_histogram(stat="count") + labs(title = paste("Distribution of ", feature)) +
    theme_minimal() + theme(plot.title = element_text(size = 8)) + theme(axis.title.x = element_text(size=8))
})

grid.arrange(grobs = plots, ncol = 4)
```
When comparing the summary statistics of the individuals who are classified as leverage points for out regression model, we find that there is quite a significant difference in their Math and Biology S11 scores and the average. In this subset of our sample, the average score on Math and Biology for the S11 exams was 73.09 and 72.04 respectively. In the sample as a whole, the average Math and biology scores for the S11 exams were 64 and 63.87 respectively.
This could potentially mean that these individuals were driving up the coefficient on Math and Biology S11 in our regression, explaining why they were significantly higher the other S11 test coefficients despite having similar distributions and correlations with the outcome variable G_SC.

## Influential Points
```{r}
# Influential Points
# 620 across all methods
inflm.SR <- influence.measures(model5)
influential_indices <- which(apply(inflm.SR$is.inf, 1, any))

#influential_df <- as.data.frame(inflm.SR$is.inf)
#influential_df <- influential_df[apply(influential_df, 1, any), ]
#nrow(influential_df)
influential_df <- df_encoded[influential_indices, ]
influential_df
```

## Checking whether to remove problematic observations
```{r}
# Combining all three
all_bad_indices <- unique(c(outlier_indices, leverage_indices, influential_indices))
length(all_bad_indices)
```

## Model Revised: Removing all problematic observations
```{r}
# Revised model without the bad points
model_revised <- lm(G_SC ~ MAT_S11 + ENG_S11 + CR_S11 + CC_S11 + BIO_S11 + SEL_IHE + OCC_FATHER + I(MAT_S11*BIO_S11), data = df_encoded[-all_bad_indices, ])
summary(model_revised)
anova(model_revised)
plot_diagnostics(model_revised, data = df_encoded[-all_bad_indices, ])
deviance(model_revised)
```

### Bro idk what ur really doing here figure it out
```{r}
model_temp <- lm(G_SC ~ SEL_IHE, data = df_encoded[-all_bad_indices, ])
model_epic <- feature_selection(model_temp, features, 4000, c("SEL_IHE"), data = df_encoded[-all_bad_indices, ])

summary(model_epic)
anova(model_epic)
plot_diagnostics(model_epic, data = df_encoded[-all_bad_indices, ])
```

```{r}
model_temp <- lm(G_SC ~ SEL_IHE, data = df_encoded[-all_bad_indices, ])
model_epic <- feature_selection(model_temp, features, 4000, c("SEL_IHE"), data = df_encoded[-all_bad_indices, ])

summary(model_epic)
anova(model_epic)
plot_diagnostics(model_epic, data = df_encoded[-all_bad_indices, ])
```

## Model Out: Removing only Outliers
```{r}
model_out <- lm(G_SC ~ MAT_S11 + ENG_S11 + CR_S11 + CC_S11 + BIO_S11 + SEL_IHE + OCC_FATHER + I(MAT_S11*BIO_S11), data=df_encoded[-outlier_indices, ])
summary(model_out)
anova(model_out)
plot_diagnostics(model_out, data=df_encoded[-outlier_indices, ])
deviance(model_out)
vif(model_out)
```

## Model Inf: Removing only Influential Points
```{r}
model_inf <- lm(G_SC ~ MAT_S11 + ENG_S11 + CR_S11 + CC_S11 + BIO_S11 + SEL_IHE + OCC_FATHER + I(MAT_S11*BIO_S11), data=df_encoded[-influential_indices, ])
summary(model_inf)
anova(model_inf)
plot_diagnostics(model_inf, data=df_encoded[-influential_indices, ])
deviance(model_inf)
vif(model_inf)
```

Although removing influential points improves adjusted R^2 by a significant
amount, we will not be removing them. This is because they make up a large
percentage of the dataset, (5.5%) and so they represent a valid subset of the 
true population. Almost all leverage points are influential points, so keeping 
influential points is the same as keeping leverage points.

Outliers, however, represent special cases and consist of only 3 observations.
Therefore, we will be removing them.

The final model then is model_out. Reiterating this here as a conclusion.
```{r}
model_final <- lm(G_SC ~ MAT_S11 + ENG_S11 + CR_S11 + CC_S11 + BIO_S11 + SEL_IHE + OCC_FATHER + I(MAT_S11*BIO_S11), data=df_encoded[-outlier_indices, ])
summary(model_final)
anova(model_final)
plot_diagnostics(model_final, data=df_encoded[-outlier_indices, ])
deviance(model_final)
vif(model_final)
```








