---
title: "Analysis"
output: pdf_document
---

```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library(corrplot)
library(gridExtra)
library(car)
```


# Data Cleaning and Preprocessing
```{r}
df <- read.csv("data/data.csv")
head(df)
```

```{r}
# Adding AVG_S11
df <- df %>% 
  mutate("AVG_S11" = (MAT_S11 + CR_S11 + CC_S11 + BIO_S11 + ENG_S11) / 5)

# Making continuous columns into factors
continuous_cols <- c("MAT_S11", "CR_S11", "CC_S11", "BIO_S11", "ENG_S11", "G_SC", "AVG_S11",
                     "QR_PRO", "CR_PRO", "CC_PRO", "ENG_PRO", "WC_PRO", "FEP_PRO", "G_SC")

df <- df%>%
  mutate(across(!all_of(continuous_cols), as.factor))

# Checking missing observations
print("Missing Observations in Each Column")
colSums(is.na(df))

# removing identifies and other uninformative variables relating
df <- df %>% select(-c(...10, X, COD_S11, Cod_SPro, QUARTILE, X2ND_DECILE, PERCENTILE))
```

```{r}
# Separating Numeric and Categorical
df_numeric <- df %>% select(all_of(continuous_cols))
df_categorical <- df %>% select(-c(continuous_cols))

# Summary Statistics
summary(df_numeric)
summary(df_categorical) # frequency of all values
```

## Cleaning and Processing: Categorical Data
```{r}
# Making each value in each category unique 
lapply(df_categorical, function(x) head(unique(x), 15))

# Getting number of levels for each categorical variable
df_categorical %>% summarise_all(~ n_distinct(.))

# Getting number of missing values (encoded as 0's)
df_categorical %>%
  summarise_all(~ sum(. == 0))
```

```{r}
# since not too many missing values, we can just drop them
df <- df %>%
  filter(EDU_FATHER != 0, EDU_MOTHER != 0, OCC_FATHER != 0, OCC_MOTHER != 0, 
         STRATUM != 0, SISBEN != 0, REVENUE != 0, JOB != 0)

# selecting again for future but with the cleaned data
df_numeric <- df %>% select(all_of(continuous_cols))
df_categorical <- df %>% select(-c(continuous_cols))
```

## Categorical Features: Encodings and Transformations
```{r}
# Label Encoding Categorical Variables as numbers
# For ordinal Categories (EDU_FATHER, EDU_MOTHER, PEOPLE_HOUSE, REVENUE, STRATUM, SISBEN)
# SEL and SEL_IHE are already encoded as ordinal

ordinal_features <- list(
  EDU_FATHER = c("Not sure", "Ninguno", "Incomplete primary", "Complete primary", 
                 "Incomplete Secundary", "Complete Secundary", 
                 "Incomplete technical or technological", "Complete technique or technology",
                 "Incomplete Professional Education", "Complete professional education",
                 "Postgraduate education"),
  EDU_MOTHER = c("Not sure", "Ninguno", "Incomplete primary", "Complete primary", 
                 "Incomplete Secundary", "Complete Secundary", 
                 "Incomplete technical or technological", "Complete technique or technology",
                 "Incomplete Professional Education", "Complete professional education",
                 "Postgraduate education"),
  PEOPLE_HOUSE = c("One", "Two", "Three", "Four", "Five", "Six", "Seven", "Eight", "Nueve",
                   "Ten", "Once", "Twelve or more"),
  STRATUM = c("Stratum 1", "Stratum 2", "Stratum 3", "Stratum 4", "Stratum 5", "Stratum 6"),
  SISBEN = c("It is not classified by the SISBEN", "Level 1", "Level 2", "Level 3",
             "Esta clasificada en otro Level del SISBEN"),
  JOB = c("No", "Yes, 20 hours or more per week", "Yes, less than 20 hours per week")
)

df_cat_encoded <- data.frame(df_categorical)
df_cat_encoded[names(ordinal_features)] <- lapply(names(ordinal_features), function(feature){
  df_cat_encoded[[feature]] <- as.factor(as.integer(factor(df_categorical[[feature]], levels = ordinal_features[[feature]], ordered = TRUE)))
})
```

```{r}
# special ordinal encoding for Revenue and people_house since can tell 
# value of difference from one category to another
# revenue level is average of LMMW
df_cat_encoded <- df_cat_encoded %>%
  mutate(REVENUE = case_when(
    REVENUE == "less than 1 LMMW" ~ 0.5,
    REVENUE == "Between 1 and less than 2 LMMW" ~ 1.5,
    REVENUE == "Between 2 and less than 3 LMMW" ~ 2.5, 
    REVENUE == "Between 3 and less than 5 LMMW" ~ 4, 
    REVENUE == "Between 5 and less than 7 LMMW" ~ 6, 
    REVENUE == "Between 7 and less than 10 LMMW" ~ 8.5,
    REVENUE == "10 or more LMMW" ~ 10
  ))

df_cat_encoded$PEOPLE_HOUSE <- as.numeric(df_cat_encoded$PEOPLE_HOUSE)
```

```{r}
# Encoding All the Yes No Variables
# Yes no variables have inconsistent encoding in original dataset (ie some have Yes = 1 No = 0, others have No = 1, Yes = 0)
yes_no_encoding <- c("No", "Yes") # encodes all no's as 0 and yes as 1
yes_no_features <- c("INTERNET", "TV", "COMPUTER", "WASHING_MCH", "MIC_OVEN", 
                        "CAR", "FRESH", "PHONE", "MOBILE", "DVD")
df_cat_encoded[yes_no_features] <- lapply(yes_no_features, function(feature){
  df_cat_encoded[[feature]] <- as.factor(as.integer(factor(df_categorical[[feature]], levels = yes_no_encoding)) - 1)
})
```

```{r}
# gender encoding unique for interpretation, make it yes no for male
df_cat_encoded$GENDER <- as.factor(as.integer(factor(df_categorical$GENDER)) - 1)
names(df_cat_encoded)[names(df_cat_encoded) == "GENDER"] <- "GENDER_MALE"
```

```{r}
# encoding everything else
remaining_features <- c("SCHOOL_NAME", "SCHOOL_NAT", "SCHOOL_TYPE", "UNIVERSITY",
                        "ACADEMIC_PROGRAM", "SEL", "SEL_IHE", "OCC_FATHER", "OCC_MOTHER")

df_cat_encoded[remaining_features] <- lapply(remaining_features, function(feature){
  df_cat_encoded[[feature]] <- as.factor(as.integer(factor(df_categorical[[feature]])))
})
```

```{r}
# final category encoded dataframe
head(df_cat_encoded)
```

## Categorical Variables: Visualizations
```{r}
# visualizing categorical features with less than 12 unique observations per category
categorical_features <- names(df_cat_encoded)
df_cat_lt_12 <- df_cat_encoded[sapply(df_cat_encoded[categorical_features], function(x) length(unique(x)) < 12)]

plots <- lapply(names(df_cat_lt_12)[0:16], function(feature) {
  ggplot(df_cat_encoded, aes_string(x = feature)) + geom_histogram(stat="count") + labs(title = paste("Distribution of ", feature)) +
    theme_minimal() + theme(plot.title = element_text(size = 8)) + theme(axis.title.x = element_text(size=8))
})

grid.arrange(grobs = plots, ncol = 4)

plots <- lapply(names(df_cat_lt_12)[16:length(names(df_cat_lt_12))], function(feature) {
  ggplot(df_cat_encoded, aes_string(x = feature)) + geom_histogram(stat="count") + labs(title = paste("Distribution of ", feature)) +
    theme_minimal() + theme(plot.title = element_text(size = 8)) + theme(axis.title.x = element_text(size=8))
})

grid.arrange(grobs = plots, ncol = 4)
```

```{r}
# Plots for All Remaining Categorical Features
df_cat_gt_12 <- df_cat_encoded[sapply(df_cat_encoded[categorical_features], function(x) length(unique(x)) >= 12)]

for (feature in names(df_cat_gt_12)) {
  print(
    ggplot(df_cat_encoded, aes_string(x = feature)) + geom_histogram(stat="count") + 
    labs(title = paste("Distribution of ", feature)) + theme_minimal() + 
    theme(plot.title = element_text(size = 8)) + theme(axis.title.x = element_text(size=8))
  )
}

# grid.arrange(grobs = plots, ncol = 4)
```

## Numerical Variables: Visualizations
```{r}
# Plots to show distributions of each variable
plots <- lapply(names(df_numeric), function(feature) {
  ggplot(df_numeric, aes_string(x = feature)) + geom_histogram(bins = 30) + labs(title = paste("Distribution of ", feature)) +
    theme_minimal() + theme(plot.title = element_text(size = 8)) + theme(axis.title.x = element_text(size=8))
})

grid.arrange(grobs = plots, ncol = 4)
```

```{r}
# dropping features that are from same test period as G_SC to improve clarity in visualization
numeric_to_drop <- c("QR_PRO", "CR_PRO", "CC_PRO", "ENG_PRO", "WC_PRO", "FEP_PRO")
df_numeric_filtered <- df_numeric %>% select(-all_of(numeric_to_drop))
```


```{r}
# Scatterplots
plot(df_numeric_filtered)
```


```{r}
# Correlation matrix heatmap
corr_matrix <- cor(df_numeric_filtered)
corrplot(corr_matrix, method = "color", tl.srt = 45, addCoef.col = "white", number.cex = 0.8)
```

## Numerical and Categorical Visualizations
```{r}
# Boxplots for each numerical and categorical feature with less than 13 unique values
for (cat_feature in names(df_cat_lt_12)) {
  plots <- lapply(names(df_numeric), function(numeric_feature) {
    ggplot(df_encoded, aes_string(x = cat_feature, y = numeric_feature)) +
    geom_boxplot() +
    labs(title = paste(numeric_feature, " and ", cat_feature), y = numeric_feature,
         x = cat_feature) +
    theme_minimal() +
    theme(plot.title = element_text(size = 8)) +
    theme(axis.title.x = element_text(size=8))
  })

  grid.arrange(grobs = plots, ncol = 4)
}
```

## #Fully Cleaned and Processed Dataset
```{r}
df_encoded <- cbind(df_cat_encoded, df_numeric)
```

### Removing Unnecessary Features
AVG_S11 dropped as it is purely meant to get a sense of the average scores on the Saber 11 tests. By including this in the model, we may introduce perfect multicolinearity, breaking the assumptions of linear regression. Using AVG_S11 instead of the individual S11 tests will also reduce reduce the amount of information contained within the model, reducing predictive power and interpretability.

The scores on the PRO tests were dropped as our outcome G_SC is derived from such scores.

Cateogrical Variables with over 20+ categories were dropped due to large imbalances in the categories and lack of information provided by the categories. This only includes the variables of "University", "School Name", and "Academic Program"

```{r}
features <- c(names(df_numeric), names(df_cat_encoded))
features <- features[!features %in% c(
  "QR_PRO", "CR_PRO", "CC_PRO", "ENG_PRO", "WC_PRO", "FEP_PRO",
  "G_SC", "SCHOOL_NAME", "UNIVERSITY", "ACADEMIC_PROGRAM", "AVG_S11"
  )]
```

# Model Building and Selection

## Function for Model Selection

### Feature Selection

As our dataset contains a large set of predictors and we also consider all possible interaction terms, using built in functions for feature selection creates excess strain on computing and often does not work properly. To compensate for this, we chose to make our own algorithm for forward selection that seeks to minimize RSS. But we customize our function with a threshold to prevent overfitting and reduce computing time.

```{r}
# feature selection function
feature_selection <- function(model, features_list, threshold, excluded = c(), data = df_encoded) {
  
  features_not_in_model <- setdiff(features_list, excluded)
  rss_old <- deviance(model)
  rss_reduction_threshold <- threshold 
  rss_reduction <- Inf
  model <- model
  while (rss_reduction >= rss_reduction_threshold) {
    rss_values <- numeric(length(features_not_in_model))
    for (i in 1:length(features_not_in_model)) {
      new_formula <- as.formula(paste(Reduce(paste, deparse(formula(model))), " + ", features_not_in_model[i]))
      model_temp <- lm(new_formula, data = data)
      rss_new <- deviance(model_temp)
      rss_reduction_temp <- rss_old - rss_new
      rss_values[i] <- rss_reduction_temp  
    }
    max_rss_ind <- which.max(rss_values)
    new_formula <- as.formula(paste(Reduce(paste, deparse(formula(model))), " + ", features_not_in_model[max_rss_ind]))
    model <- lm(new_formula, data = data)
    rss_reduction <- max(rss_values)
    rss_old <- deviance(model)
    features_not_in_model <- features_not_in_model[-max_rss_ind]
    print(rss_reduction)
  }
  return(model)
}
```

```{r}
# function for finding the first predictor to add
feature_selection_first <- function(model, feature_list, df = df_encoded) {
  rss_values <- numeric(length(feature_list))
  # Getting first ele
  for (i in 1:length(feature_list)){
    new_formula <- as.formula(paste(Reduce(paste, deparse(formula(model))), " + ", features_not_in_model[i]))
    model_temp <- lm(new_formula, data = df)
    rss_new <- deviance(model_temp)
    rss_values[i] <- rss_new
  }
  min_rss_ind <- which.min(rss_values)
  print(min_rss_ind)
  print(feature_list[i])
  new_formula <- as.formula(paste(Reduce(paste, deparse(formula(model))), " + ", features_not_in_model[min_rss_ind]))
  model <- lm(new_formula, data = df)
  return(model)
}
```

```{r}
# model diagnostics function to create residual plots
plot_diagnostics <- function(model, data = df_encoded) {
 res <- residuals(model)
  y_pred <- fitted(model)
  plot(y_pred, res, abline(h = 0, lty = 2, col = "red"))
  
  std_residual <- rstandard(model)
  hist(std_residual, breaks = 50, xlab="Standardized Residual", main="Histogram of Standardized Residuals")
  
  qqnorm(res)
  
  y_pred <- fitted(model)
  plot(y_pred, data$G_SC)

}
```

```{r}
# creating list of all features with interaction terms
interactions <- c()
for (i in 1:(length(features)-1)) {
  for (j in (i+1):length(features)) {
    interaction <- paste(c(features[i], features[j]), collapse = ":")
    interactions <- append(interactions, interaction)
  }
}

features_with_int <- c(features, interactions)
```

# Model Building and Model Selection

## Model Prev: Preliminary Model 
```{r}
# preliminary model for comparison
model_prev <- lm(G_SC~ AVG_S11 + SEL+ SEL_IHE + GENDER_MALE + SCHOOL_NAT, data=df_encoded)
summary(model_prev)
anova(model_prev)
plot_diagnostics(model_prev)
```

## Model 0: Baseline Model
```{r}
# baseline model with all features
formula <- as.formula(paste("G_SC ~", paste(features, collapse = " + ")))
model0 <- lm(as.formula(formula), data = df_encoded)

summary(model0)
vif(model0)                # assessing multicolinearity
plot_diagnostics(model0)   # diagnostics for baseline model
```

## Model 1: Naive Feature Selection

Trying feature selection with only raw variables, no transformations,
no interactions.

```{r}
# SEL_IHE was selected by feature_selection_first
model1 <- lm("G_SC ~ SEL_IHE", data = df_encoded)  
model1 <- feature_selection(model1, features, 4000, c("SEL_IHE"))

summary(model1)
anova(model1)
vif(model1)
deviance(model1)
plot_diagnostics(model1)
```

## Model 2: Just Interactions

Trying feature selection with list of all features including interactions.
```{r}
model2 <- lm("G_SC ~ SEL_IHE", data = df_encoded)
model2 <- feature_selection(model2, features_with_int, 80000, c("SEL_IHE"))

summary(model2)
anova(model2)
plot_diagnostics(model2)
```

## Model 3: Adding Interactions to Model 1
```{r}
# taking original model and adding interactions
model1_preds <- attr(terms(model1), "term.labels")
model3 <- feature_selection(model1, features_with_int, 30000, excluded=model1_preds)

summary(model3)
anova(model3)
plot_diagnostics(model3)
```

## Final Model Contenders
```{r}
# best model without interactions
model4 <- lm(G_SC ~ MAT_S11 + ENG_S11 + CR_S11 + CC_S11 + BIO_S11 + SEL_IHE + OCC_FATHER, data=df_encoded)
summary(model4)
anova(model4)
plot_diagnostics(model4)
deviance(model4)
vif(model4)
```

```{r}
# best model with interactions
model5 <- lm(G_SC ~ MAT_S11 + ENG_S11 + CR_S11 + CC_S11 + BIO_S11 + SEL_IHE + OCC_FATHER + I(MAT_S11*BIO_S11), data=df_encoded)
summary(model5)
anova(model5)
plot_diagnostics(model5)
deviance(model5)
vif(model5)
```

We chose to go ahead with model 5 as the additional interaction term provides a slightly higher adjusted R^2, slightly reduced RSS and increased significance to our research question.

```{r}
# transfering data into .dta format for stata regression tables
library(haven)

df$REVENUE <- as.numeric(df_encoded$REVENUE)

write_dta(df, "df_stata.dta")
```

# Investigating Problematic Observations

## Outliers
```{r}
# 3 outliers
stanres <- rstudent(model5)
sum(stanres >= 4)
outlier_indices <- which(stanres >= 4)

df_outlier <- df[outlier_indices, ]
df_outlier
```
All individuals who are outliers did really poorly on S11 (avg of ~40%) but did significantly better on G_SC (avg of ~60%).
- Score for the PRO exams on average were just a lot higher than on S11 exams. Although can't see from the summary of the numeric data, from visualizations, the pro tests are a lot more left skewed. But should still be noted that G_SC and FEP_PRO are fairly normally distributed.
- Mostly likely caused by unknown factors not representative of the population as a whole.

## Leverage Points
```{r}
# Leverage Points
# 114 leverage points
leverage_values <- hatvalues(model5)
p <- length(coef(model5)) - 1
n <- nrow(df_encoded) 
leverage_threshold <- 3*(p+1)/n       # not using 2*(p+1) / n, being more strict

leverage_indices <- which(leverage_values > leverage_threshold)
df_leverage <- df_encoded[leverage_indices, ]
head(df_leverage)
```

### Visualizations for further testing LPs
```{r}
# Visualizations of Numeric Data for Leverage Points
plots <- lapply(names(df_numeric), function(feature) {
  ggplot(df_leverage, aes_string(x = feature)) + geom_histogram(bins = 30) + labs(title = paste("Distribution of ", feature)) +
    theme_minimal() + theme(plot.title = element_text(size = 8)) + theme(axis.title.x = element_text(size=8))
})

grid.arrange(grobs = plots, ncol = 4)
```

```{r}
# visualizations for categorical features of leverage points data
plots <- lapply(names(df_cat_lt_12)[0:16], function(feature) {
  ggplot(df_leverage, aes_string(x = feature)) + geom_histogram(stat="count") + labs(title = paste("Distribution of ", feature)) +
    theme_minimal() + theme(plot.title = element_text(size = 8)) + theme(axis.title.x = element_text(size=8))
})

grid.arrange(grobs = plots, ncol = 4)

plots <- lapply(names(df_cat_lt_12)[16:length(names(df_cat_lt_12))], function(feature) {
  ggplot(df_leverage, aes_string(x = feature)) + geom_histogram(stat="count") + labs(title = paste("Distribution of ", feature)) +
    theme_minimal() + theme(plot.title = element_text(size = 8)) + theme(axis.title.x = element_text(size=8))
})

grid.arrange(grobs = plots, ncol = 4)
```
When comparing the summary statistics of the individuals who are classified as leverage points for out regression model, we find that there is quite a significant difference in their Math and Biology S11 scores and the average. In this subset of our sample, the average score on Math and Biology for the S11 exams was 73.09 and 72.04 respectively. In the sample as a whole, the average Math and biology scores for the S11 exams were 64 and 63.87 respectively.
This could potentially mean that these individuals were driving up the coefficient on Math and Biology S11 in our regression, explaining why they were significantly higher the other S11 test coefficients despite having similar distributions and correlations with the outcome variable G_SC.
The individuals represented by the leverage points largely had fathers that were stay at home (OCC_FATHER = HOME (level 4)). This is interesting as it could mean that there is a very large difference between individuals with fathers who work versus fathers who stay at home. The distribution for the occupation of the mother is also interesting as it is very bimodal with the majority of mothers either staying at home or working a technical/professional job (level 11).

## Influential Points
```{r}
# Influential Points
# 620 across all methods
inflm.SR <- influence.measures(model5)
influential_indices <- which(apply(inflm.SR$is.inf, 1, any))

influential_df <- df_encoded[influential_indices, ]
influential_df
```

## Checking whether to remove problematic observations
```{r}
# Combining all three
all_bad_indices <- unique(c(outlier_indices, leverage_indices, influential_indices))
length(all_bad_indices)
```

## Model Revised: Removing all problematic observations
```{r}
# Revised model without the bad points
model_revised <- lm(G_SC ~ MAT_S11 + ENG_S11 + CR_S11 + CC_S11 + BIO_S11 + SEL_IHE + OCC_FATHER + I(MAT_S11*BIO_S11), data = df_encoded[-all_bad_indices, ])
summary(model_revised)
anova(model_revised)
plot_diagnostics(model_revised, data = df_encoded[-all_bad_indices, ])
deviance(model_revised)
```

## Model Out: Removing only Outliers
```{r}
model_out <- lm(G_SC ~ MAT_S11 + ENG_S11 + CR_S11 + CC_S11 + BIO_S11 + SEL_IHE + OCC_FATHER + I(MAT_S11*BIO_S11), data=df_encoded[-outlier_indices, ])
summary(model_out)
anova(model_out)
plot_diagnostics(model_out, data=df_encoded[-outlier_indices, ])
deviance(model_out)
vif(model_out)
```

## Model Inf: Removing only Influential Points
```{r}
model_inf <- lm(G_SC ~ MAT_S11 + ENG_S11 + CR_S11 + CC_S11 + BIO_S11 + SEL_IHE + OCC_FATHER + I(MAT_S11*BIO_S11), data=df_encoded[-influential_indices, ])
summary(model_inf)
anova(model_inf)
plot_diagnostics(model_inf, data=df_encoded[-influential_indices, ])
deviance(model_inf)
vif(model_inf)
```

Although removing influential points improves adjusted R^2 by a significant
amount and results in significantly better diagnostic plots, we will not be 
removing them. This is because they make up a sizeable percentage of the dataset, 
(5.5%) and so they represent a valid subset of the true population. Almost all 
leverage points are influential points, so keeping influential points is the 
same as keeping leverage points.

Outliers, however, represent special cases and consist of only 3 observations.
Therefore, we will be removing them.

The final model then is model_out. Reiterating this here as a conclusion.
```{r}
model_final <- lm(G_SC ~ MAT_S11 + ENG_S11 + CR_S11 + CC_S11 + BIO_S11 + SEL_IHE + OCC_FATHER + I(MAT_S11*BIO_S11), data=df_encoded[-outlier_indices, ])
summary(model_final)
anova(model_final)
plot_diagnostics(model_final, data=df_encoded[-outlier_indices, ])
deviance(model_final)
vif(model_final)
```
```{r}
# transfering data to stata for regression tables
write_dta(df[-outlier_indices, ], "df_stata_no_outlier.dta")
```
