---
title: "Analysis"
output: pdf_document
---

```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library(corrplot)
library(gridExtra)
```


## Data Preprocessing
```{r}
df <- read.csv("data/data.csv")
head(df)
```


```{r}
df <- df %>% 
  mutate("AVG_S11" = (MAT_S11 + CR_S11 + CC_S11 + BIO_S11 + ENG_S11) / 5)

continuous_cols <- c("MAT_S11", "CR_S11", "CC_S11", "BIO_S11", "ENG_S11", "G_SC", "AVG_S11",
                     "QR_PRO", "CR_PRO", "CC_PRO", "ENG_PRO", "WC_PRO", "FEP_PRO", "PERCENTILE",
                     "G_SC")

df <- df%>%
  mutate(across(!all_of(continuous_cols), as.factor))

print("Missing Observations in Each Column")
colSums(is.na(df))
```

```{r}
# removing id and other variables
df <- df %>% select(-c(...10, X, COD_S11, Cod_SPro))
```

```{r}
df_numeric <- df %>% select(all_of(continuous_cols))
df_categorical <- df %>% select(-c(continuous_cols))
```

```{r}
summary(df_numeric)
```

```{r}
summary(df_categorical)
```
```{r}
lapply(df_categorical, function(x) head(unique(x), 15))
```

```{r}
# number of distinct observations in each column for categorical
df_categorical %>% summarise_all(~ n_distinct(.))
```
* NOTE (TO BE REMOVED LATER): Should probably do multiple analyses, with some having the varaibles that have lots of observations (20+) as fixed effects rather than variables we care for

```{r}
# missing values were encoded as 0's
df_categorical %>%
  summarise_all(~ sum(. == 0))
```

```{r}
# since not too many missing values, can just drop them
df <- df %>%
  filter(EDU_FATHER != 0, EDU_MOTHER != 0, OCC_FATHER != 0, OCC_MOTHER != 0, 
         STRATUM != 0, SISBEN != 0, REVENUE != 0, JOB != 0)
```

```{r}
# 2 encodings for 1 person in house (One and Once)
df$PEOPLE_HOUSE[df$PEOPLE_HOUSE=="Once"] <- "One"
```

```{r}
# selecting again but with the cleaned data
df_numeric <- df %>% select(all_of(continuous_cols))
df_categorical <- df %>% select(-c(continuous_cols))
```

## Visualizations of Numerical Variables
```{r}
# distributions of each variable
plots <- lapply(names(df_numeric), function(feature) {
  ggplot(df_numeric, aes_string(x = feature)) + geom_histogram(bins = 30) + labs(title = paste("Distribution of ", feature)) +
    theme_minimal() + theme(plot.title = element_text(size = 8)) + theme(axis.title.x = element_text(size=8))
})

grid.arrange(grobs = plots, ncol = 4)
```

## Categorical Feature Things
```{r}
##TODO: MAKE THIS CODE PRETTIER CUZ HOLY FRICK THIS IS SO UGLY

# investigating the values for ordinal data
# levels(factor(df_categorical$EDU_FATHER))
# levels(factor(df_categorical$EDU_MOTHER))
# levels(factor(df_categorical$PEOPLE_HOUSE))
# levels(factor(df_categorical$REVENUE))
# levels(factor(df_categorical$STRATUM))
# levels(factor(df_categorical$SISBEN))
# levels(factor(df_categorical$SEL))
# levels(factor(df_categorical$SEL_IHE))
# levels(factor(df_categorical$MIC_OVEN))
```

```{r}
# Label Encoding Categorical Variables as numbers

# For ordinal Categories (EDU_FATHER, EDU_MOTHER, PEOPLE_HOUSE, REVENUE, STRATUM, SISBEN)
# SEL and SEL_IHE are already encoded as ordinal
ordinal_features <- list(
  EDU_FATHER = c("Not sure", "Ninguno", "Incomplete primary", "Complete primary", 
                 "Incomplete Secundary", "Complete Secundary", 
                 "Incomplete technical or technological", "Complete technique or technology",
                 "Incomplete Professional Education", "Complete professional education",
                 "Postgraduate education"),
  EDU_MOTHER = c("Not sure", "Ninguno", "Incomplete primary", "Complete primary", 
                 "Incomplete Secundary", "Complete Secundary", 
                 "Incomplete technical or technological", "Complete technique or technology",
                 "Incomplete Professional Education", "Complete professional education",
                 "Postgraduate education"),
  PEOPLE_HOUSE = c("One", "Two", "Three", "Four", "Five", "Six", "Seven", "Eight", "Nueve",
                   "Ten", "Once", "Twelve or more"),
  REVENUE = c("less than 1 LMMW", "Between 1 and less than 2 LMMW",
              "Between 2 and less than 3 LMMW", "Between 3 and less than 5 LMMW", 
              "Between 5 and less than 7 LMMW", "Between 7 and less than 10 LMMW",
              "10 or more LMMW"),
  STRATUM = c("Stratum 1", "Stratum 2", "Stratum 3", "Stratum 4", "Stratum 5", "Stratum 6"),
  SISBEN = c("It is not classified by the SISBEN", "Level 1", "Level 2", "Level 3",
             "Esta clasificada en otro Level del SISBEN"),
  JOB = c("No", "Yes, 20 hours or more per week", "Yes, less than 20 hours per week")
)

df_cat_encoded <- data.frame(df_categorical)
df_cat_encoded[names(ordinal_features)] <- lapply(names(ordinal_features), function(feature){
  df_cat_encoded[[feature]] <- as.factor(as.integer(factor(df_categorical[[feature]], levels = ordinal_features[[feature]], ordered = TRUE)))
})
```

```{r}
# Encoding All the Yes No Variables
# Yes no variables have inconsistent encoding in original dataset (ie some have Yes = 1 No = 0, others have No = 1, Yes = 0)
yes_no_encoding <- c("No", "Yes") # encodes all no's as 0 and yes as 1
yes_no_features <- c("INTERNET", "TV", "COMPUTER", "WASHING_MCH", "MIC_OVEN", 
                        "CAR", "FRESH", "PHONE", "MOBILE", "DVD")
df_cat_encoded[yes_no_features] <- lapply(yes_no_features, function(feature){
  df_cat_encoded[[feature]] <- as.factor(as.integer(factor(df_categorical[[feature]], levels = yes_no_encoding)) - 1)
})
```

```{r}
# gender encoding unique for interpretation, make it yes no for male
df_cat_encoded$GENDER <- as.factor(as.integer(factor(df_categorical$GENDER)) - 1)
names(df_cat_encoded)[names(df_cat_encoded) == "GENDER"] <- "GENDER_MALE"
```

```{r}
# encoding everything else
remaining_features <- c("SCHOOL_NAME", "SCHOOL_NAT", "SCHOOL_TYPE", "UNIVERSITY",
                        "ACADEMIC_PROGRAM", "SEL", "SEL_IHE", "X2ND_DECILE", "QUARTILE",
                        "OCC_FATHER", "OCC_MOTHER")

df_cat_encoded[remaining_features] <- lapply(remaining_features, function(feature){
  df_cat_encoded[[feature]] <- as.factor(as.integer(factor(df_categorical[[feature]])))
})
```

```{r}
# TODO: MAKE THIS CODE A LOT MORE ELEGENT CUZ THIS IS SUPER UGLY
# View Encodings for everything else
levels(factor(df_categorical$GENDER))
# levels(factor(df_categorical$SCHOOL_NAME))
levels(factor(df_categorical$SCHOOL_NAT))
levels(factor(df_categorical$SCHOOL_TYPE))
# levels(factor(df_categorical$UNIVERSITY))
# levels(factor(df_categorical$ACADEMIC_PROGRAM))
levels(factor(df_categorical$SEL))
levels(factor(df_categorical$SEL_IHE))
# levels(factor(df_categorical$OCC_FATHER))
# levels(factor(df_categorical$OCC_MOTHER))
```

```{r}
head(df_cat_encoded)
```
```{r}
# visualizing categorical features with less than 12 unique observations per category
categorical_features <- names(df_cat_encoded)
df_cat_lt_12 <- df_cat_encoded[sapply(df_cat_encoded[categorical_features], function(x) length(unique(x)) < 12)]

plots <- lapply(names(df_cat_lt_12)[0:16], function(feature) {
  ggplot(df_cat_encoded, aes_string(x = feature)) + geom_histogram(stat="count") + labs(title = paste("Distribution of ", feature)) +
    theme_minimal() + theme(plot.title = element_text(size = 8)) + theme(axis.title.x = element_text(size=8))
})

grid.arrange(grobs = plots, ncol = 4)

plots <- lapply(names(df_cat_lt_12)[16:length(names(df_cat_lt_12))], function(feature) {
  ggplot(df_cat_encoded, aes_string(x = feature)) + geom_histogram(stat="count") + labs(title = paste("Distribution of ", feature)) +
    theme_minimal() + theme(plot.title = element_text(size = 8)) + theme(axis.title.x = element_text(size=8))
})

grid.arrange(grobs = plots, ncol = 4)
```

```{r}
# TODO: FIX ENCODINGS FOR THIS GARBAGE
# Plots for All Remaining Categorical Features
df_cat_gt_12 <- df_cat_encoded[sapply(df_cat_encoded[categorical_features], function(x) length(unique(x)) >= 12)]

for (feature in names(df_cat_gt_12)) {
  print(
    ggplot(df_cat_encoded, aes_string(x = feature)) + geom_histogram(stat="count") + 
    labs(title = paste("Distribution of ", feature)) + theme_minimal() + 
    theme(plot.title = element_text(size = 8)) + theme(axis.title.x = element_text(size=8))
  )
}

# grid.arrange(grobs = plots, ncol = 4)
```
```{r}
df_encoded <- cbind(df_cat_encoded, df_numeric)
```


# Exploratory Data Analysis
## Numerical Features Only

```{r}
# dropping features that cause multicolinearity
numeric_to_drop <- c("QR_PRO", "CR_PRO", "CC_PRO", "ENG_PRO", "WC_PRO", "FEP_PRO", "PERCENTILE")
df_numeric_filtered <- df_numeric %>% select(-all_of(numeric_to_drop))
```


```{r}
# Scatterplots
plot(df_numeric_filtered)
```


```{r}
# Correlation matrix heatmap
library(corrplot)

corr_matrix <- cor(df_numeric_filtered)
corrplot(corr_matrix, method = "color", tl.srt = 45, addCoef.col = "white", number.cex = 0.8)
```

## Numerical and Categorical
```{r}
# Boxplots for each numerical and categorical feature with less than 13 unique values
for (cat_feature in names(df_cat_lt_12)) {
  plots <- lapply(names(df_numeric), function(numeric_feature) {
    ggplot(df_encoded, aes_string(x = cat_feature, y = numeric_feature)) + 
    geom_boxplot() + 
    labs(title = paste(numeric_feature, " and ", cat_feature), y = numeric_feature, 
         x = cat_feature) + 
    theme_minimal() + 
    theme(plot.title = element_text(size = 8)) + 
    theme(axis.title.x = element_text(size=8))
  })
  
  grid.arrange(grobs = plots, ncol = 4)
}
```

# Initial Model
```{r}
# TODO: CLEAN UP ALL CODE AFTER THIS
# AVG_S11 dropped to prevent perfect multicolinearity
# PRO tests dropped since G_SC calculated from PRO tests
# Other indicators such as decile and percentile dropped as not informative
# Any categorical variable with 20+ category dropped cuz yeah
features <- c(names(df_numeric), names(df_cat_encoded))
features <- features[!features %in% c(
  "QR_PRO", "CR_PRO", "CC_PRO", "ENG_PRO", "WC_PRO", "FEP_PRO",
  # "MAT_S11", "CR_S11", "CC_S11", "BIO_S11", "ENG_S11",
  "PERCENTILE", "G_SC", "SCHOOL_NAME", "UNIVERSITY", 
  "ACADEMIC_PROGRAM", "X2ND_DECILE", "QUARTILE",
  "AVG_S11"
  )]
```

```{r}
# baseline model
formula <- as.formula(paste("G_SC ~", paste(features, collapse = " + ")))
model0 <- lm(as.formula(formula), data = df_encoded)

summary(model0)
```


```{r}
# assessing multicolinearity
library(car)

vif(model0)
```


```{r}
# model diagnostics function
plot_diagnostics <- function(model) {
 res <- residuals(model)
  y_pred <- fitted(model)
  plot(y_pred, res, abline(h = 0, lty = 2, col = "red"))
  
  std_residual <- rstandard(model)
  hist(std_residual, breaks = 50, xlab="Standardized Residual", main="Histogram of Standardized Residuals")
  
  qqnorm(res) 
}
```

```{r}
# diagnostics for baseline model
plot_diagnostics(model0)
```

# Feature Selection
```{r}
# due to high multicolinearity between parent education, the average parent education will be used instead
df_encoded["AVG_PARENT_EDU"] <- as.factor(ceiling((as.numeric(df_encoded$EDU_FATHER) + as.numeric(df_encoded$EDU_MOTHER))/2))
```

```{r}
# feature_selection
feature_selection <- function(model, features_list, threshold) {
  
  features_not_in_model <- features_list[features_list != "SEL_IHE"]
  # print(features_not_in_model)
  rss_old <- deviance(model)
  rss_reduction_threshold <- threshold # arbitrary value, previously was 20000
  rss_reduction <- Inf
  model <- model
  while (rss_reduction >= rss_reduction_threshold) {
    #i made this size of features list so we can index instead of appending 
    rss_values <- numeric(length(features_not_in_model))
    for (i in 1:length(features_not_in_model)) {
      new_formula <- as.formula(paste(Reduce(paste, deparse(formula(model))), " + ", features_not_in_model[i]))
      model_temp <- lm(new_formula, data = df_encoded)
      rss_new <- deviance(model_temp)
      rss_reduction_temp <- rss_old - rss_new
      rss_values[i] <- rss_reduction_temp  
      # made this index instead of append, also ensures index of rss_values n features_not_in_model match
    }
    max_rss_ind <- which.max(rss_values)
    new_formula <- as.formula(paste(Reduce(paste, deparse(formula(model))), " + ", features_not_in_model[max_rss_ind]))
    model <- lm(new_formula, data= df_encoded)
    rss_reduction <- max(rss_values)
    rss_old <- deviance(model)
    features_not_in_model <- features_not_in_model[-max_rss_ind]
    print(rss_reduction)
  }
  return(model)
}
```

```{r}
# function for finding the first predictor to add
feature_selection_first <- function(model, feature_list) {
  rss_values <- numeric(length(feature_list))
  # Getting first ele
  for (i in 1:length(feature_list)){
    new_formula <- as.formula(paste(Reduce(paste, deparse(formula(model))), " + ", features_not_in_model[i]))
    model_temp <- lm(new_formula, data = df_encoded)
    rss_new <- deviance(model_temp)
    rss_values[i] <- rss_new
  }
  min_rss_ind <- which.min(rss_values)
  print(min_rss_ind)
  print(feature_list[i])
  new_formula <- as.formula(paste(Reduce(paste, deparse(formula(model))), " + ", features_not_in_model[min_rss_ind]))
  model <- lm(new_formula, data = df_encoded)
  return(model)
}
```


```{r}
model1 <- lm("G_SC ~ SEL_IHE", data = df_encoded)  # changed this to sel_ihe

model1 <- feature_selection(model1, features, 4000)
```

```{r}
summary(model1)
anova(model1)
vif(model1)
deviance(model1)
```

```{r}
# diagnostics for second model
plot_diagnostics(model1)
```
# Transformations and Interactions
```{r}
# adding interaction terms
interactions <- c()
for (i in 1:(length(features)-1)) {
  for (j in (i+1):length(features)) {
    interaction <- paste(c(features[i], features[j]), collapse = ":")
    interactions <- append(interactions, interaction)
  }
}

```

```{r}
features_with_int <- c(features, interactions)
features_with_int
```


```{r}
start.time <- Sys.time() # for funsies to see the O(n^2) runtime
model2 <- lm("G_SC ~ SEL_IHE", data = df_encoded)
model2 <- feature_selection(model2, features_with_int, 80000)
end.time <- Sys.time()
time.taken <- end.time - start.time
time.taken
```


```{r}
summary(model2)
anova(model2)
```


```{r}
plot_diagnostics(model2)
```
```{r}
model3 <- feature_selection(model1, features_with_int, 30000)

summary(model3)
anova(model3)
plot_diagnostics(model3)
```


```{r}
# preliminary model for comparison
model_prev <- lm(G_SC~ AVG_S11 + SEL+ SEL_IHE + GENDER_MALE + SCHOOL_NAT, data=df_encoded)
summary(model_prev)
anova(model_prev)
plot_diagnostics(model_prev)
```

```{r}
# model including both
model3 <- lm(G_SC ~ MAT_S11 + ENG_S11 + CR_S11 + CC_S11 + BIO_S11 + SEL_IHE + REVENUE + OCC_FATHER, data=df_encoded)
summary(model3)
anova(model3)
plot_diagnostics(model3)
deviance(model3)
vif(model3)
```




